Appendix 3: Estimating species interactions

### 2015-10-25 Harris_2015/manuscript-materials/draft.Rmd
various

**Title:** Estimating species interactions from observational data with Markov
networks

@@

**Abstract:** Estimating species interactions from observational data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, the negative correlation between
two competing species can be reversed in the presence of a third species that is
capable of outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can predict
the direct and indirect consequences of any possible species interaction matrix.
Interactions in these models can be estimated from observational data via
maximum likelihood. Using simulated landscapes with known pairwise interaction
strengths, I evaluated Markov networks and six existing approaches. The
Markov networks consistently outperformed other methods, correctly isolating
direct interactions between species pairs even when indirect interactions
largely overpowered them. Two computationally efficient approximations, based on
linear and generalized linear models, also performed well. Indirect effects
reliably caused a common null modeling approach to produce incorrect inferences,
however.

@@

To the extent that nontrophic species interactions (such as competition) affect
community assembly, ecologists might expect to find signatures of these interactions in
species composition data [@macarthur_population_1958; @diamond_island_1975].
Despite decades of work and several major controversies, however
[@lewin_santa_1983; @strong_ecological_1984; @gotelli_swap_2003;
@connor_checkered_2013], existing methods for detecting competition's effects
on community structure are unreliable [@gotelli_empirical_2009]. In particular,
species' effects on one another can become lost in the complex web of direct and
indirect interactions in real assemblages. For example, the competitive
interaction between the two shrub species in Figure 1A can become obscured by
their shared tendency to occur in unshaded areas (Figure 1B). While ecologists
have long known that indirect effects can overwhelm direct ones at the landscape
level [@dodson_complementary_1970; @levine_competitive_1976], the vast majority of our methods for drawing
inferenes from observational data do not control for these effects
[e.g. @diamond_island_1975; @strong_ecological_1984; @gotelli_empirical_2009;
@veech_probabilistic_2013; @pollock_understanding_2014]. To the extent that
indirect interactions like those in Figure 1 are generally important
[@dodson_complementary_1970], existing methods will thus not generally provide
much evidence regarding species' direct effects on one another. The goal of this
paper is to resolve this long-standing problem.

While competition doesn't reliably reduce co-occurrence rates at the whole-landscape
level (as most of our methods assume), it nevertheless does leave a signal in
the data (Figure 1C). Specifically, after partitioning the data set into shaded
sites and unshaded sites, there will be co-occurrence deficits in each subset
that might not be apparent at the landscape level. More generally,
we can obtain much better estimates of the association between two species from
their conditional relationships (i.e. by controlling for other species in the
network) than we could get from their overall co-occurrence rates. This kind of
precision is difficult to obtain from null models, which begin with the
assumption that all the pairwise interactions are zero and thus don’t need to
be controlled for.  Nevertheless, null models have dominated this field for more
than three decades [@strong_ecological_1984; @gotelli_empirical_2009].

Following recent work by @azaele_inferring_2010 and @fort_statistical_2013, this
paper shows that Markov networks [undirected graphical models also known as
Markov random fields; @murphy_machine_2012] can provide a framework for
understanding the landscape-level consequences of pairwise species interactions,
and for detecting them from observed presence-absence matrices. Markov networks
have been used in many scientific fields in similar contexts for decades, from
physics [where nearby particles interact magnetically; @cipra_introduction_1987]
to spatial statistics [where adjacent grid cells have correlated values;
@harris_contact_1974; @gelfand_modelling_2005]. While community ecologists
explored some related approaches in the 1980's [@whittam_species_1981],
they used severe approximations that led to unintelligible results
[e.g. "probabilities" greater than one; @gilpin_factors_1982].

Below, I introduce Markov networks and show how they can be used to simulate
landscape-level data or to make exact predictions about the direct and
indirect consequences of possible interaction matrices. Then, using simulated
data sets where the "true" interactions are known, I compare this approach with
several existing methods. Finally, I discuss opportunities for extending the
approach presented here to other problems in community ecology, e.g. quantifying
the overall effect of species interactions on occurrence rates
[@roughgarden_competition_1983] and disentangling the effects of biotic versus
abiotic interactions on species composition [@kissling_towards_2012;
@pollock_understanding_2014].

@@

#### Markov networks.
Markov networks provide a framework for translating back and forth between the
conditional relationships among species (Figure 1C) and the kinds of species
assemblages that these relationships produce. Here, I show how a set of
conditional relationships can be used to determine how groups of species
can co-occur. Methods for estimating conditional relationships from data are
discussed in the next section.

@@

$p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{i\neq j}\beta_{ij}y_i y_j).$

@@

Here, $\alpha_{i}$ is an intercept term determining the amount that the presence
of species $i$ contributes to the log-probability of $\vec{y}$; it directly
controls the prevalence of species $i$. Similarly, $\beta_{ij}$ is the amount
that the co-occurrence of species $i$ and species $j$ contributes to the
log-probability; it controls the conditional relationship between two species,
i.e. the probability that they will be found together, after controlling for the
other species in the network (Figure 2A, Figure 2B). For example, $\beta_{ij}$
might have a value of $+2$ for two mutualists, indicating that the odds of
observing one species are $e^2$ times higher in sites where its partner is
present than in comparable sites where its partner is absent. Because the
relative probability of a presence-absence vector increases when
positively-associated species co-occur and decreases when negatively-associated
species co-occur, the model tends---all else equal---to produce assemblages that
have many pairs of positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect).

Of course, if all else is *not* equal (e.g. Figure 1, where the presence of one
competitor is associated with release from another competitor), then species'
marginal association rates can differ from this expectation. Determining the
marginal relationships between species from their conditional interactions
entails summing over the different possible assemblages (Figure 2B). This
becomes intractable when the number of possible assemblages is large, though
several methods beyond the scope of this paper can be employed to keep the
calculations feasible [@lee_learning_2012; @salakhutdinov_learning_2008].
Alternatively, as noted below, some common linear and generalized linear methods
can also be used as computationally efficient approximations to the full network
[@lee_learning_2012; @loh_structure_2013].

#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a matrix of ones and zeros
indicating which species are present at which sites). When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for all
of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will always converge on the global maximum. This
maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
that would be expected to produce exactly the observed co-occurrence frequencies
on average [i.e. maximizing the likelihood matches the sufficient statistics of the model distribution
to the sufficient statistics of the data; @murphy_machine_2012]. I used the
rosalia package [@harris_rosalia_2015] for the R programming language
[@r_core_team_r_2015] to optimize the Markov network parameters. The package was
named after Santa Rosalia, the patron saint of biodiversity, whose supposedly
miraculous healing powers played an important rhetorical role in the null model
debates of the 1970's and 1980's [@lewin_santa_1983].

#### Simulated landscapes.
In order to compare different methods, I simulated two sets of landscapes using
known parameters. The first set included the three competing species shown in
Figure 1. For each of 1000 replicates, I generated a landscape with 100 sites by
sampling from a probability distribution defined by the figure's interaction
coefficients (Appendix 1). Each of the methods described below was then
evaluated on its ability to correctly infer that the two shrub species competed
with one another, despite their frequent co-occurrence.

I also simulated a second set of landscapes using a stochastic community model
based on generalized Lotka-Volterra dynamics, as described in Appendix 2. In
these simulations, each species pair was randomly assigned to either compete for
a portion of the available carrying capacity (negative interaction) or to act as
mutualists (positive interaction).  Here, mutualisms operate by mitigating the
effects of intraspecific competition on each partner's death rate.  For these
analyses, I simulated landscapes with up to 20 species and 25, 200, or 1600
sites (50 replicates per landscape size; see Appendix 2).

#### Recovering species interactions from simulated data.
<!-- exp(qlogis(.975, scale = 2)) -->

I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package [@harris_rosalia_2015] to fit Markov newtork
models,  as described above. For the analyses with 20 species, I added a very
weak logistic prior distribution on the $\alpha$ and $\beta$ terms with scale 2
to ensure that the model estimates were always finite. The bias introduced by
this prior should be small: the 95% credible interval on $\beta$ only requires
that one species' effect on the odds of observing a different species to be less
than a factor of 1500 (which is not much of a constraint). The logistic
distribution was chosen because it is convex and has a similar shape to the
Laplace distribution used in LASSO regularization (especially in the tails), but
unlike the Laplace distribution it is differentiable everywhere and does not
force any estimates to be exactly zero. To confirm that this procedure produced
stable estimates, I compared its estimates on 50 bootstrap replicates
(Appendix 4).

I also evaluated six alternative methods: five from the existing literature,
plus a novel combination of two of these methods. The first alternative interaction
metric was the sample correlation between species' presence-absence vectors,
which summarizes their marginal association.  Next, I used partial correlations,
which summarize species' conditional relationships [@albrecht_spatial_2001;
@faisal_inferring_2010]. In the context of non-Gaussian data, the partial
correlation can be thought of as a computationally efficient approximation to
the full Markov network model [@loh_structure_2013]. This sort of model is very
common for estimating relationships among genes and gene products
[@friedman_sparse_2008]. Because partial correlations are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
`pcor.shrink` function with the default settings [@schafer_corpcor_2014].

The third alternative, generalized linear models (GLMs), can also be
thought of as a computationally efficient approximation to the Markov
network [@lee_learning_2012].  Following @faisal_inferring_2010,  I fit
regularized logistic regression models [@gelman_weakly_2008] for each species,
using the other species on the landscape as predictors. To avoid the
identifiability problems associated with directed cyclic graphs
[@schmidt_modeling_2012], I then symmetrized the relationships within species
pairs via averaging.

@@

I then evaluated the relationship between each method's estimates and the "true"
interaction strengths among all of the species pairs from the larger simulated
landscapes. This determined which of the methods provide a consistent way to know
how strong species interactions are---regardless of which species were present
in a particular data set or how many observations were taken. Because the
different methods mostly describe species interactions on different scales
(e.g. correlations versus $Z$ scores versus regression coefficients), I used
linear regression through the origin to rescale the different estimates produced
by each method so that they had a consistent interpretation. After rescaling
each method’s estimates, I calculated squared errors between the scaled
interaction estimates and “true” interaction values across all the simulated
data sets. These squared errors determined the proportion of variance explained
for different combinations of model type and landscape size (compared with a
null model that assumed all interaction strengths to be zero).

@@

#### Twenty species.
Despite some variability across contexts (Figure 3A), the four methods that
controlled for indirect effects clearly performed the best: the Markov network
explained the largest portion of the variance in the "true" interaction
coefficients (35% overall), followed by the generalized linear models (30%),
partial correlations from the raw presence-absence data (28%), and partial
correlations from BayesComm, the joint species distribution model (26%).  The
benefit of choosing the full Markov network over the other three methods was
largest on the smaller landscapes, which are also the ones that are most
representative of typical analyses in this field [@gotelli_empirical_2009].

The three methods that did not attempt to control for indirect interactions all
explained less than 20% of the variance. Of these, the sample correlation matrix
based on the raw data performed the best (19%), followed by the null model
(15%) and BayesComm's correlation matrix (11%). Although these last three
methods had different $R^2$ values, there was a close mapping among
their estimates (especially after controlling for the size of the simulated
landscapes; Figure 3B).  This suggests that the effect sizes from the null model
(and, to a lesser extent, the correlation matrices from joint species
distribution models) only contain noisy versions of the same information that
could be obtained more easily and interpretably by calculating correlation
coefficients between species' presence-absence vectors.

Bootstrap resampling indicated that the above ranking of the different methods
was robust (Appendix 3). In particular, the 95% confidence interval of the
bootstrap distribution indicated that the Markov network's overall $R^2$ value
was between 14 and 18 percent higher than the second-most effective
method (generalized linear models) and between 2.12 and 2.38 times higher than
could be achieved by the null model (Pairs).  Bootstrap resampling of a 200-site
landscape also confirmed that the rosalia package's estimates of species'
conditional relationships were robust to sampling variation for reasonably-sized
landscapes (Appendix 4).

@@

The results presented above show that Markov networks can reliably recover
species' pairwise interactions from observational data, even for cases where
a common null modeling technique reliably fails. Specifically, Markov networks
were successful even when direct interactions were largely overwhelmed by
indirect effects (Figure 1). For cases where fitting a Markov network is
computationally infeasible, these results also indicate that partial covariances
and generalized linear models (the two methods that estimated conditional
relationships rather than marginal ones) can both provide useful approximations.
The partial correlations' success on simulated data may not carry over to real
data sets, however; @loh_structure_2013 show that the linear approximations can
be less reliable in cases where the true interaction matrix contains more
structure (e.g. guilds or trophic levels). Similarly, the approximation involved
in using separate generalized linear models for each species can occasionally
lead to catastrophic overfitting with small-to-moderate sample sizes
[@lee_learning_2012]. For these reasons, it will usually be best to fit a Markov
network rather than one of the alternative methods when one’s computational
resources allow it.

@@

Markov networks---particularly the Ising model for binary networks---have been
studied for nearly a century [@cipra_introduction_1987], and the models'
properties, capabilities, and limits are well-understood in a
huge range of applications. Using the same framework for species interactions
would thus allow ecologists to tap into an enormous set of existing
discoveries and techniques for dealing with indirect effects, stability, and
alternative stable states. Numerous other extensions are possible: for example,
the states of the interaction network can be modeled as a function of
the local abiotic environment [@lee_learning_2012], which would provide a
rigorous and straightforward approach to the difficult and important task of
incorporating whole networks of biotic interactions into species distribution
models [@kissling_towards_2012; @pollock_understanding_2014], leading to a
better understanding of the interplay between biotic and abiotic effects on
community structure. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow one species to affect the sign or
strength of the relationship between two other species, tipping the balance
between facilitation and exploitation [@bruno_inclusion_2003].

@@

>>>>

**Title:** Estimating species interactions from co-occurrence data with Markov
networks

@@

**Abstract:** Estimating species interactions from co-occurrence data is one of
the most controversial tasks in community ecology. One difficulty is that a
single pairwise interaction can ripple through an ecological network and produce
surprising indirect consequences. For example, the negative correlation between
two competing species can be reversed in the presence of a third species that is
capable of outcompeting both of them. Here, I apply models from statistical
physics, called Markov networks or Markov random fields, that can predict
the direct and indirect consequences of any possible species interaction matrix.
Interactions in these models can also be estimated from observed co-occurrence
rates via maximum likelihood. Using simulated landscapes with known pairwise
interaction strengths, I evaluated Markov networks and six existing approaches.
The Markov networks consistently outperformed other methods, correctly isolating
direct interactions between species pairs even when indirect interactions
largely overpowered them. Two computationally efficient approximations, based on
controlling for indirect effects with linear or generalized linear models,
also performed well. Indirect effects reliably caused a common null modeling
approach to produce incorrect inferences, however.

@@

To the extent that nontrophic species interactions (such as competition) affect
community assembly, ecologists might expect to find signatures of these interactions in
species composition data [@macarthur_population_1958; @diamond_island_1975].
Despite decades of work and several major controversies, however
[@lewin_santa_1983; @strong_ecological_1984;
@connor_checkered_2013], existing methods for detecting competition's effects
on community structure are unreliable [@gotelli_empirical_2009]. In particular,
species' effects on one another can become lost in the complex web of
indirect effects. For example, the competitive interaction between the two shrub
species in Figure 1A can become obscured by their shared tendency to occur in
unshaded areas (Figure 1B). While ecologists have long known that indirect
effects can overwhelm direct ones at the landscape level
[@dodson_complementary_1970; @levine_competitive_1976], the vast majority of our
methods for drawing inferences from observational data do not control for these effects
[e.g. @diamond_island_1975; @strong_ecological_1984; @gotelli_empirical_2009;
@veech_probabilistic_2013; @pollock_understanding_2014]. To the extent that
indirect interactions like those in Figure 1 are generally important
[@dodson_complementary_1970], existing methods will thus not generally provide
much evidence regarding species' direct effects on one another. The goal of this
paper is to resolve this long-standing problem.

While competition doesn't reliably reduce co-occurrence rates at the whole-landscape
level (as most methods assume), it nevertheless does leave a signal in
the data (Figure 1C). For example, after partitioning the data set into shaded
sites and unshaded sites, there will be co-occurrence deficits in each subset
that might not be apparent at the whole-landscape level. More generally,
controlling for other species in the network will often be important for obtaining
reliable estimates of direct (conditional, or all-else-equal) effects.  This kind of precision is
difficult to obtain from null models, which only simulate the most extreme
possible scenario: that *all* direct and indirect interactions are exactly zero.
Nevertheless, null models have dominated this  field for more than three decades
[@strong_ecological_1984; @gotelli_empirical_2009].

Following @azaele_inferring_2010, this paper shows that Markov networks
[undirected graphical models also known as
Markov random fields; @murphy_machine_2012] can provide a framework for
understanding the landscape-level consequences of pairwise species interactions,
and for detecting them from observed presence-absence matrices. Markov networks
have been used in many scientific fields in similar contexts for decades, from
physics [where nearby particles interact magnetically; @cipra_introduction_1987]
to spatial statistics [where adjacent grid cells have correlated values;
@harris_contact_1974; @gelfand_modelling_2005]. While community ecologists
explored some related approaches in the 1980's [@whittam_species_1981],
they used severe approximations that led to unintelligible results
[e.g. "probabilities" greater than one; @gilpin_factors_1982].

Below, I introduce Markov networks and show how they can be used to simulate
landscape-level data from an interaction matrix or to make exact predictions
about the direct and indirect consequences of possible interaction matrices.
Then, using simulated data sets where the "true" interactions are known,
I compare this approach with several existing methods. Finally, I discuss
opportunities for extending the approach presented here to other problems in
community ecology, e.g. quantifying the overall effect of species interactions
on occurrence rates [@roughgarden_competition_1983] and disentangling the
effects of biotic versus abiotic interactions on species composition
[@pollock_understanding_2014].

@@

#### Markov networks.
Markov networks provide a framework for translating back and forth between the
conditional (all-else-equal) relationships among species (Figure 1C) and the
kinds of species assemblages that these relationships produce. Here, I show how
a set of conditional relationships can be used to determine how groups of species
can co-occur. Methods for estimating conditional relationships from data are
discussed in the next section.

@@

$p(\vec{y}; \alpha, \beta) \propto exp(\sum_{i}\alpha_i y_i + \sum_{<ij>}\beta_{ij}y_i y_j),$

@@

\noindent where the second sum is over all $\binom{n}{2}$ species pairs. In this model,  $\alpha_{i}$ is
an intercept term determining the amount that the presence
of species $i$ contributes to the log-probability of $\vec{y}$; it directly
controls the prevalence of species $i$. Similarly, $\beta_{ij}$ is the amount
that the co-occurrence of species $i$ and species $j$ contributes to the
log-probability; it controls the conditional relationship between two species,
i.e. the probability that they will be found together, after controlling for the
other species in the network (Figure 2A, Figure 2B). For example, $\beta_{ij}$
might have a value of $+2$ for two mutualists, indicating that the odds of
observing one species are $e^2$ times higher in sites where its partner is
present than in comparable sites where its partner is absent. Because the
relative probability of a presence-absence vector increases when
positively-associated species co-occur and decreases when negatively-associated
species co-occur, the model tends---all else equal---to produce assemblages that
have many pairs of positively-associated species and relatively few pairs of
negatively-associated species (exactly as an ecologist might expect).

Of course, if all else is *not* equal (e.g. Figure 1, where the presence of one
competitor is associated with release from another competitor), then species'
overall co-occurrence patterns can be more complicated. Determining the
marginal relationships between species from their conditional interactions
entails summing over the different possible assemblages (Figure 2B). This
becomes intractable when the number of possible assemblages is large, though
several methods beyond the scope of this paper can be employed to keep the
calculations feasible [@lee_learning_2012; @salakhutdinov_learning_2008].
Alternatively, as noted below, some common linear and generalized linear methods
can also be used as computationally efficient approximations to the full network
[@lee_learning_2012; @loh_structure_2013].

#### Estimating $\alpha$ and $\beta$ coefficients from presence-absence data.
In the previous section, the values of $\alpha$ and $\beta$ were known and the
goal was to make predictions about possible species assemblages. In
practice, however, ecologists will often need to estimate the parameters from
an observed co-occurrence matrix (i.e. from a set of independent $\vec{y}$ vectors
indicating which species are present at each site on the landscape). When the number of species
is reasonably small, one can compute exact maximum likelihood estimates for all
of the $\alpha$ and $\beta$ coefficients given a presence-absence matrix by
optimizing $p(\vec{y}; \alpha, \beta)$. Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces [@murphy_machine_2012],
ensuring that this procedure will always converge on the global maximum. This
maximum represents the unique combination of $\alpha$ and $\beta$ coefficients
that would be expected to produce exactly the observed co-occurrence frequencies
on average [i.e. maximizing the likelihood ensures that the sufficient
statistics of the model distribution match the sufficient statistics of the data;
@murphy_machine_2012]. I used the rosalia package [@harris_rosalia_2015] for the
R programming language [@r_core_team_r_2015] to optimize the Markov network
parameters. The package was named after Santa Rosalia, the patron saint of
biodiversity, whose mythos played an important rhetorical role in the null model
debates of the 1970's and 1980's [@lewin_santa_1983].

#### Simulated landscapes.
I simulated several sets of landscapes using known parameters as a test bed for
different statistical methods. The first set of landscapes included the three
competing species shown in Figure 1. For each of 1000 replicates, I generated a
landscape with 100 sites by sampling from a probability distribution defined by
the figure's interaction coefficients (Appendix 1). Each of the methods
described below was then evaluated on its ability to correctly infer that the
two shrub species competed with one another, despite their frequent co-occurrence.

I then generated landscapes with up to 20 interacting species at 25, 200, or
1600 sites using three increasingly complex models (50 replicates for each
combination of size and model). As described in Appendix [[X]], I randomly drew
the “true” coefficient values for each replicate landscape in these simulations
so that most species pairs interacted negligibly, a few pairs interacted very
strongly, and competition was three times more common than facilitation.

The first set of 20-species landscapes, like the landscapes with three species,
were generated directly from a Markov network to ensure that the model could
recover the parameters used to generate the "observed" co-occurrence data.
Then, I added two environmental factors that varied from location to location
across the simulated landscapes, and simulated a new set of co-occurrence data
so that species' $\alpha$ coefficients depended on these environmental variables.
The latter set of simulated landscapes provide an important test of the methods’
ability to distinguish co-occurrence patterns that were generated from pairwise
biotic interactions from those that were generated by external forces like
abiotic environmental filtering. This task was made especially difficult
because---as with most analyses of presence-absence data for co-occurrence
patterns---the inference procedure did not have access to any information about
the environmental or spatial variables that helped shape the landscape [cf
@connor_checkered_2013].  Finally, I simulated the population dynamics of 20
interacting species, as described in Appendix [[X]]

#### Recovering species interactions from simulated data.

I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package [@harris_rosalia_2015] to fit Markov network
models,  as described above. For the analyses with 20 species, I added a weak
regularizer (equivalent to a logistic prior with scale 2) to ensure that the
model estimates were always finite [[Appendix]].

I also evaluated six alternative methods: five from the existing literature,
plus a novel combination of two of these methods. The first alternative interaction
metric was the sample correlation between species' presence-absence vectors,
which summarizes their marginal association.  Next, I used partial correlations,
which summarize species' conditional relationships. This approach is
common in molecular biology [@friedman_sparse_2008], but less so in ecology
[see @albrecht_spatial_2001 and @faisal_inferring_2010 for two exceptions].
In the context of non-Gaussian data, the partial correlation can be thought of
as a computationally efficient approximation to the full Markov network model
[@loh_structure_2013]. Because partial correlations are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor package’s
`pcor.shrink` function with the default settings [@schafer_corpcor_2014].

The third alternative, generalized linear models (GLMs), also provide a
computationally efficient approximation to the Markov network
[@lee_learning_2012].  Following @faisal_inferring_2010,  I fit regularized
logistic regression models [@gelman_weakly_2008] for each species, using the
other species on the landscape as predictors. This produced two interaction
estimates for each species pair; one for the effect of species $i$ on species
$j$ and one for the reverse. These two coefficients are not identifiable from
the data, however [@schmidt_modeling_2012], so I used their average as an
overall measure of the overall relationship between species $i$ and species $j$.

@@

I then evaluated the relationship between each method's estimates and the "true"
interaction strengths among all of the species pairs from the larger simulated
landscapes. This determined which of the methods provide a consistent way to
know how strong species interactions are---regardless of which species were
present in a particular data set or how many observations were taken. To ensure
that the different test statistics (e.g. correlations versus $Z$ scores) were on
a common scale, I rescaled them using linear regression through the origin. After
rescaling each method’s estimates, I calculated squared errors between the scaled
interaction estimates and “true” interaction values across the simulated data
sets. These squared errors determined the proportion of variance explained
for different combinations of model type and landscape size (compared with a
null model that assumed all interaction strengths to be zero).

@@

#### Twenty species.
Despite some variability across contexts (Figure 3), the four methods that
controlled for indirect effects (the Markov network, the generalized linear
models, and the two partial correlation-based methods) clearly outperformed those
that did not.  Of these, the Markov network consistently performed the best,
especially on the smaller landscapes [which are also the ones that are most
representative of typical analyses in this field; @gotelli_empirical_2009].

Of the methods that did not control for indirect effects, Figure 3 shows that
simple correlation coefficients provided a more reliable indicator of species'
true interaction strengths than either the joint species distribution model
(BayesComm) or the null model (Pairs). Interestingly, the estimates from these
three approaches were tightly correlated [[Figure 3B]]. This suggests that the
effect sizes from the null model (and, to a lesser extent, the correlation
matrices from joint species distribution models) only contain noisy versions of
the same information that could be obtained more easily and interpretably by
calculating correlation coefficients between species' presence-absence vectors.

Bootstrap resampling indicated that the above ranking of the different methods
was robust (Appendix 3). In particular, the 95% confidence interval of the
bootstrap distribution indicated that the Markov network's overall $R^2$ value
was between [[14 and 18]] percent higher than the second-most effective
method (generalized linear models) and between [[2.12 and 2.38]] times higher than
could be achieved by the null model (Pairs).

@@

The results presented above show that Markov networks can reliably recover
species' pairwise interactions from observational data, even for cases where
environmental heterogeneity and indirect interactions cause ecologists' typical
null modeling approaches to reliably fail. For cases where fitting a Markov network is
computationally infeasible, these results also indicate that partial covariances
and generalized linear models can both provide useful approximations.
The partial correlations' success on simulated data may not carry over to real
data sets, however; @loh_structure_2013 show that the linear approximations can
be less reliable in cases where the true interaction matrix contains more
structure (e.g. guilds or trophic levels). Similarly, the approximation involved
in using separate generalized linear models for each species can occasionally
lead to catastrophic overfitting with small-to-moderate sample sizes
[@lee_learning_2012]. For these reasons, it will usually be best to fit a Markov
network rather than one of the alternative methods when one’s computational
resources allow it.

@@

Markov networks---particularly the Ising model for binary networks---have been
studied for nearly a century [@cipra_introduction_1987], and the models'
properties, capabilities, and limits are well-understood in a
huge range of applications. Using the same framework for species interactions
would thus allow ecologists to tap into an enormous set of existing
discoveries and techniques for dealing with indirect effects, stability, and
alternative stable states. Numerous other extensions are possible: for example,
the states of the interaction network can be modeled as a function of
the local abiotic environment [@lee_learning_2012], which would provide a
rigorous and straightforward approach to the difficult and important task of
incorporating whole networks of biotic interactions into species distribution
models [@pollock_understanding_2014], leading to a
better understanding of the interplay between biotic and abiotic effects on
community structure. There are even methods [@whittam_species_1981;
@tjelmeland_markov_1998] that would allow one species to affect the sign or
strength of the relationship between two other species, tipping the balance
between facilitation and exploitation [@bruno_inclusion_2003].

@@

**Figure 1: A.** A small network of three competing species. The tree (top) tends not to
co-occur with either of the two shrub species, as indicated by the strongly
negative coefficient linking them. The two shrub species also compete with one
another, but more weakly (circled coefficient). **B.** In spite of the
competitive interactions between the two shrub species, their shared tendency to
occur in locations without trees makes their occurrence vectors positively
correlated (circled). **C.** Controlling for the tree species' presence with a
conditional (all-else-equal) approach such as a partial covariance or a
Markov network leads to correct identification of the negative shrub-shrub
interaction (circled).

**Figure 2: A.** A small Markov network with two species, defined by its $\alpha$ and
$\beta$ values.  The abiotic environment favors the occurrence of each species
($\alpha >0$), particularly species 2 ($\alpha_2 > \alpha_1$). The negative
$\beta$ coefficient linking these two species implies that they co-occur less
than expected under independence. **B.** These coefficients determine the
probabilities of all four possible presence-absence combinations for Species 1
and Species 2. $\alpha_1$ is added to the exponent whenever Species 1 is present
($y_1 = 1$), but not when it is absent ($y_1 = 0$).  Similarly, the exponent
includes $\alpha_2$ only when species $2$ is present ($y_2 = 1$), and
$\beta$ only when both are present ($y_1y_2 = 1$). The normalizing constant $Z$,
ensures that the four probabilities sum to 1.  In this case, $Z$ is
about 18.5.  **C.** The expected frequencies of all possible
co-occurrence patterns between the two species of interest, as calculated in the
previous panel. **D.** If, as assumed in null model simulations, $\beta_{12}$
equaled zero (e.g. if the species no longer competed for the same resources),
then the reduction in competition would allow each species to increase its
overall occurrence rate.

**Figure 3: A.** Proportion of variance in interaction coefficients explained by each
method versus number of sampled locations.
**B.** The $Z$-scores produced by the null model ("Pairs") for each pair of
species can be predicted using the correlation between the presence-absence
vectors of those same species and from the number of sites on the landscape.
